---
title: |
  <center> R script Assignment 1</center>
  <center> Maximum Likelyhood Estimation </center>
  
author: "Jan Izquierdo, Eloi Vilella"
date: "2023-10-20"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1

#### ML estimation of a one-parameter distribution. Let X be a random variable with probability density f(x\|B)=Be-Bx+1 with x\>1/B. We consider a random sample of n observations of this distribution.

#### *1.a)*Write down the likelihood function for a sample of n observations of this distribution

$L(B|x)=∏^n_{i=1}B*e^{-Bx+1}$

#### *1.b)* Obtain the log-likelihood function

We solve using logarithmic properties



Setting libraries and some variables

```{r setting the scene}
library("ggplot2")
db<-read.table("./Assignment_data/A1_MLE.dat")
m<-mean(db$V1)
sd1<-sd(db$V1)
variance<-var(db$V1)
```

#### ***1.e)*** Download the file Sample.dat, which contains sample of observations from this probability distribution.Determine the sample size and calculate the value of the ML estimator for this sample.

.

```{r, echo=TRUE}
n <- length(db$V1)
x <- db$V1
b <- seq(min(db$V1),max(db$V1),by=0.01)
L <- (b^n)*(prod(x)^(b-1))
cat("The sample size is", n)
```

This method does not work as the sample size (n) is too big, it approaches infinity. We will calculate it this other way.

```{r}
#n/sum(x) x is all values
MLE<-100000/sum(db$V1) 
#MLE=~sd1/m
cat("MLE is", MLE)
```

#### ***1.f)*** Plot the log-likelihood function, and assess graphically if your ML estimate coincides with the maximum of this function.

.

```{r, echo=TRUE}
log_likelihood<-function(m){
  log_data<-dnorm(db$V1, mean=m, sd=sd1, log=T)
  log_likelyhood<-sum(log_data)
  return(log_likelyhood)
}

beta<-seq(min(db$V1), max(db$V1), by=0.1)#min->max or 0.1->max
log_likelihood_values <- sapply(beta, log_likelihood)

plot_data <- data.frame(Beta = beta, LogLikelihood = log_likelihood_values)

ggplot(plot_data, aes(x=beta, y=log_likelihood_values))+
          geom_line()+
          geom_vline(xintercept =100000/sum(db$V1), color = "blue", linetype = "dashed")+
          xlab("Beta")+
          ylab("Logarithmic likelihood")
```

The blue line represents where the MLE stands respect the graph

#### ***1.j)*** Calculate a 95% confidence interval for parameter B, using the dataset that you have downloaded.

```{r, echo=T}
#n is sample size, MLE is maximum likelihood estimator
error <- qt(0.975, df=n-1)*sd1/(2*sqrt(n))
lower_bound <- MLE - error
upper_bound <- MLE + error
cat("Confidence interval of 95% for parameter B is","[", lower_bound, ",", upper_bound, "]")

```

#### ***1.l)*** Make a histogram of the data, using function hist, using the argument freq=FALSE. Overplot the histogram with the estimated probability density f(x\|B), using the maximum likelihood estimate. What do you observe?.

.

```{r, echo=T}
Prob_density<- function(a, beta) {
  beta * exp(-beta * a + 1)
}
hist(db$V1, freq=F, main = "Histogram Estimated Probability Density", xlab="Population")
lines(density(db$V1), col="green", lwd=2)
curve(Prob_density(x, MLE), col = "blue", lwd = 4, add = TRUE)

```

## Exercise 2

Setting libraries and variables

```{r}
o<-1#placeholder
AA<-o^2
Aa<-2*o*(1-o)
aa<-(1-o)^2
#n<-sample size
#z<-individuals AA
#x<-individuals Aa
#y<-individuals aa
```

#### ***2.c)*** Record the form of MLE for o.

.

```{r}
log_likelihood<-function(o, z ,x ,y){
  n<-z+x+y
  AA<-o^2
  Aa<-2*o*(1-o)
  aa<-(1-o)^2
  log_lik<-log(factorial(n))-(log(factorial(z))+log(factorial(x))+log(factorial(y)))+
    z*log(AA)+x*log(Aa)+y*log(aa)
  #z*log(AA)=total individuals* prob of an individual=individuals in sample
  return(-log_lik)#negative, as optimize gets minimum, we want maximum, so we use the negative of the function
}
z<-20#example data
x<-50#example data
y<-10#example data
mle_opt<-optimize(log_likelihood, interval =c(0,1), z, x, y)
mle_o<-mle_opt$minimum
cat("MLE for o is", mle_o)
```
