---
title: |
  <center> Assignment 3 </center>
  <center> Linear Regression</center>
author: "Jan Carreras & Jan Izquierdo"
date: "2023-11-6"
output: 
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### We consider a data set reporting the daily total of bike counts conducted on the Brooklyn Bridge from 01 April 2017 to 31 October 2017 (Source: NYC Open Data: Bicycle Counts for East River Bridges). The variables included in the dataset are HIGH_T; LOW_T; PRECIP and LABOR_YESNO. The number of bikes (BB_COUNT) is used as response variable in a Poisson regression. Predictors (4 variables) we consider as potentially bearing a relationship with the number of bikes are the high and low temperature (HIGH_T; LOW_T;) the precipitation (PRECIP) and the if day of the week is a working day or not (LABOR_YESNO with values 1 and 0 to indicate yes or no, respectively).

Setting libraries
```{r, echo=T, results='hide'}
#install.packages("AER")
library(readxl)
library(ggplot2)
library(AER)
library(MASS)
```

#### *1.a)* Read the dataset, you can use instruction attach the dataset for convenient access to the variables in the dataset.
.

```{r}
data<-read_excel("./assignment_data/bicyclist_data.xls")
  n<-nrow(data)
n
```

#### *1.b)* Make a barplot of the table of the possible outcomes of the response variable BB_COUNT. Calculate descriptive statistics of response BB_COUNT. Is there, at the exploratory level, evidence that the response does not follow a Poisson distribution?
.

```{r}
bb_table <- table(data$BB_COUNT)
barplot(bb_table, main="BB_count barplot")
bb_sum<-summary(data$BB_COUNT)
pois_dis<-rpois(214, mean(data$BB_COUNT))
pois_table<-table(pois_dis)
barplot(pois_table, col="red", main="Poisson comparison barplot");barplot(bb_table,col="blue",add=T)
```
```{r, echo=F}
cat("The descriptive statistics of BB_count is\n");print(bb_sum)
cat("In the BB_count barplot we can see that, the majority of BB_COUNT values have a frequency of 1, with only
a few occurrences having a frequency of 2.\n")
cat("In the barplot above we created data in a poisson distribution arrangement(red) and compare it with 
the distribution of our data(blue), we can see that the distributions are very different so we can
concludethat BB_count does not follow a Poisson regression\n")
cat("A feature of the Poisson distribution is that the mean equal to the variance. However, in our scenario,
the variance significantly exceeds the mean, indicating that our data does not follow the distribution.
To assess it, we compute the overdispersion parameter for our model.\n")
```
```{r}
model <- glm(BB_COUNT ~ HIGH_T + LOW_T + PRECIP + LABOR_YESNO,data=data, family = poisson(link = "log"))
# Get the residual deviance and the degrees of freedom
residual_dev <- model$deviance
df <- model$df.residual

# Calculate the overdispersion parameter
overdispersion <- residual_dev/df
```
```{r, echo=F}
cat("The overdispersion parameter is:", overdispersion)
```

#### *1.c)* Perform Poisson regression of the number of BB_COUNT on HIGH_T, our first model. Report the regression equation. Is there evidence for association, and if so, what kind of association?
.

```{r}
mdl_pois_reg <-glm(formula = BB_COUNT ~ HIGH_T, data = data, family = poisson)  
summary(mdl_pois_reg)
coefficients <- coef(mdl_pois_reg)

intercept <- coefficients[1]
pendent <- coefficients[2]

```
```{r, echo=F}
cat("Regression equation:\n")
cat("ln(BB_COUNT) =", intercept, "+", pendent, "* HIGH_T\n")
cat("BB_COUNT = e^(", intercept, ") * e^(", pendent,"* HIGH_T)\n\n")
cat("There is proof of a connection; when HIGH_T rises, so does BB_COUNT. According to this equation, a 1-unit
elevation in HIGH_T leads to a BB_COUNT increase of", pendent,". This demonstrates a positive linear
correlation between the variables, denoting a positive association. The positive sign of the slope signifies
that as HIGH_T increases, BB_COUNT also experiences an increase.")
```

#### *1.d)* Make a scatter plot of BB_COUNT against HIGH_T. Add the fitted regression equation to the scatter plot.
.

```{r}
# Step 1: Convert temperature to numeric
data$HIGH_T <- as.numeric(data$HIGH_T)

# Step 1: Standardize the variables
#data$BB_COUNT_standardized <- scale(data$BB_COUNT)
#data$HIGH_T_standardized <- scale(data$HIGH_T)

#Fit the linear regression model with standardized variables
model <- glm(BB_COUNT ~ HIGH_T, data = data)
#Create the scatter plot with standardized variables
plot(data$HIGH_T, data$BB_COUNT, 
     main = "Scatter Plot with Regression", 
     xlab = "HIGH_T", ylab = "BB_COUNT", pch=19)


#Add the regression line to the plot
abline(model, col = "red")

#Get the coefficients of the regression equation
coefficients <- coef(model)
intercept <- coefficients[1]
slope <- coefficients[2]

#Add the regression equation to the plot
#cat("ln(BB_COUNT) =", intercept, "+", pendent, "* HIGH_T\n")
equation <- paste("Regression Equation: BB_COUNT =",round(intercept, 2), "+", round(slope, 2), " * HIGH_T")
text(x=70, y=4460,
     equation, 
     col="blue")
```
```{r, echo=F}
cat("We can see that there exists a positive linear correlation between these variables.")
```

#### *1.e)* Estimate the null model without predictors (BB_COUNT∼1), and also plot the equation according to this model to the plot. What do you observe?
.

```{r}
null_model<- glm(BB_COUNT ~ 1, data = data)
null_plot<-plot(data$BB_COUNT, ylab= "BB_count",
     main = "Scatter Plot", pch=19)
abline(null_model, col = "red")

coefficients <- coef(null_model)
intercept <- coefficients[1]
slope <- coefficients[2]

null_equation <- paste("Equation: BB_COUNT =", round(slope, 2), " * HIGH_T +", round(intercept, 2))
text(x=110, y=4500,
     null_equation, 
     col="blue")
```
```{r, echo=F}
cat("This model is characterized by an equation that remains unaffected by any variable, essentially
representing the null model. Consequently, it produces a horizontal line, which stays consistent along the
y-axis at the mean value. Hence, the null model is depicted as a steady line that aligns with the mean.")
```

#### *1.f)* Interpret the first model by quantifying the effect of the predictor on the average of the response. Give a 95% confidence interval for the parameter representing that effect.
.

```{r}

# Calculate 95% confidence interval for the coefficient of 'High_T'
summary(mdl_pois_reg)
pendent <- summary(mdl_pois_reg)$coefficients[2, 1]
standard_error <- summary(mdl_pois_reg)$coefficients[2, 2]
z <- qnorm(0.975)

upper <- pendent + z*standard_error
lower <- pendent - z*standard_error

```
```{r, echo=F}
cat("The 95% confidence interval is: [", lower, ",",upper, "]\n" )
cat("Within this interval, it can be inferred that a rise of 1 unit in HIGH_T is associated with a BB_COUNT
change falling between",lower, "and", upper)

```

#### *1.g)* Is the value 0 inside the interval you obtained? Is the value 1 inside the interval? What is the relevance of this?
.

```{r}

```
```{r, echo=F}
cat("Neither 1 nor 0 are within this interval [", lower, ",",upper, "]. We have a relationship between the variables under study. A 0 within our interval would indicate otherwise, as it would signify the multiplication of our variable by 0 when HIGH_T increases by one unit. A 1 within our interval would mean there would be no change, so it makes sense that neither 0 nor 1 are in our interval.")
```

#### *1.h)* Is there any indication that overdispersion is a problem for you model? Justify your answer.
.

```{r}
#Overdispersion in mdl_pois_reg
summary(mdl_pois_reg)
# Residual deviance and the degrees of freedom
res_dev_1 <- mdl_pois_reg$deviance
df_1 <- mdl_pois_reg$df.residual

# Calculate the overdispersion parameter
overdispersion_1 <- res_dev_1/df_1
```
```{r, echo=F}
cat("The overdispersion parameter is:", overdispersion_1, "\n\n")
cat("We can observe that this overdispersion is nearly twice the size of the previous one. It is too large for
us to model our data using Poisson distribution, we can say that overdispersion poses problem for our model.")
```

#### *1.i)* Formally test for overdispersion using the function dispersion test of the AER package. What is your conclusion?
.

```{r}
dispersion_test<-dispersiontest(mdl_pois_reg)
dispersion_test
```
```{r, echo=F}
cat("The results from our test strongly support our previous calculation: there is overdispersion. This
indicates that the variability in the data is higher than what would be expected from a Poisson
distribution.")
```

#### *1.j)* Calculate deviance residuals according to the first model and plot these as a function of the predicted values, using a different color for each category of LABOR_YESNO. What do you observe?
.

```{r,echo=T, results='hide'}
#Calculate residues of first model
res<-resid(mdl_pois_reg)
```
```{r}
ggplot(data, aes(x=mdl_pois_reg[["linear.predictors"]], y= res, 
  color =as.factor(LABOR_YESNO))) + 
  xlab("Predicted values") + ylab("Deviance Residuals") +
  labs(color = "LABOR_YESNO") + 
  geom_point(alpha = .8, shape = 19) + 
  scale_color_manual(values = c("green", "black"))+theme_minimal()

```
```{r, echo=F}
cat("This plot is highly scattered and lacks linearity. Which indicates that the data still doesn't fit the
model.")
```

#### *1.k)* Do a Poisson regression of BB_COUNT on HIGH_T and LABOR_YESNO. Report the fitted equation. Is there evidence for any effect of the variable LABOR_YESNO? Justify your answer.
.

```{r}
mdl_pois_reg_laboral <- glm(formula = BB_COUNT ~ HIGH_T + LABOR_YESNO,
                          data = data, family = "poisson"(link="log"))
summary(mdl_pois_reg_laboral)

coef(mdl_pois_reg_laboral )
intercept <- coef(mdl_pois_reg_laboral )[1]
slope <- coef(mdl_pois_reg_laboral )[2]
labor<-coef(mdl_pois_reg_laboral )[3]
#Create the fitted 
fitted_equation <- paste("Regression Equation: BB_COUNT =",round(intercept, 2),"+",round(slope, 2), " * HIGH_T ","+", round(labor, 2), "* LABOR_YESNO ")
```
```{r, echo=F}
print(fitted_equation)
```

#### *1.l)* Make a graphic by representing the newly fitted model in a scatterplot of BB_COUNT against HIGH_T.
.

```{r}
res_lab<-resid(mdl_pois_reg_laboral)
ggplot(data, aes(x=mdl_pois_reg_laboral[["linear.predictors"]], y= res_lab, 
  color =as.factor(LABOR_YESNO))) + geom_point()+ 
  xlab("HIGH_T") + ylab("BB_COUNT") + labs(color = "LABOR_YESNO",
  title="BB_COUNT vs HIGH_T with LABOR_YESNO distinction")+
  scale_color_manual(values = c("green", "black"))+theme_minimal()
```
```{r, echo=F}

```

#### *1.m)* Is there evidence for interaction between the variables LABOR_YESNO and HIGH_T? Justify your answer. Try to make a graphical representation of the fitted model with interaction in a scatterplot of BB_COUNT against HIGH_T.
.

```{r, echo=F}
cat("It seems that when HIGH_T is around 7.75 and 8 BB_COUNT is hgher, but there seems to be no relation
between LABOR_YESNO and HIGH_T, it would make sense if there was no relation, as they are the temperature and
if the day is laborable or not, wich are things we know are not related. Nevertheless, we will prove it")
```
```{r}
plot(data$HIGH_T, data$LABOR_YESNO)
model_H_lab <- lm(formula =HIGH_T ~ LABOR_YESNO, data = data)
anova(model_H_lab, test="Chisq")
```
```{r, echo=F}
cat("In the plot we can observe that they are not related, we confirm it through an anova on a lm test of
LABOR_YESNO on HIGH_T, where we can see that they are not significant on each other, which means that they are
not related")
```

#### *1.n)* Add the variable PRECIP to the model. Is it a significant predictor? Justify your answer.
.

```{r}
#QUIN MODEL? ...
mdl_pois_reg_laboral <- glm(formula = BB_COUNT ~ HIGH_T + LABOR_YESNO,
                          data = data, family = "poisson"(link="log"))
model <- glm(BB_COUNT ~ HIGH_T, data = data)
mdl_pois_reg <-glm(formula = BB_COUNT ~ HIGH_T, data = data, family = poisson)  
#provo mdl_pois_reg_laboral
mdl_pois_reg_laboral <- glm(formula = BB_COUNT ~ HIGH_T + LABOR_YESNO + as.double(PRECIP) ,
                          data = data, family = "poisson"(link="log"))
summary(mdl_pois_reg_laboral)

```
```{r, echo=F}
cat("PRECIP is a significant predictor for BB_COUNT, along with LABOR_YESNO and HIGH_T as they all are
significant (p_value<0.05)")
```

#### *1.o)* Add the variable LOW_T to the model. Is it a significant predictor? Justify your answer.
.

```{r}
#Mateix model q anterior, mateix error? ...
mdl_pois_reg_laboral <- glm(formula = BB_COUNT ~ HIGH_T + LABOR_YESNO + as.double(PRECIP) + as.double(LOW_T) ,
                          data = data, family = "poisson"(link="log"))
summary(mdl_pois_reg_laboral)
```
```{r, echo=F}
cat("LOW_T is a significant predictor for BB_COUNT, along with LABOR_YESNO, HIGH_T and PRECIP as they all are
significant (p_value<0.05)")
```

#### *1.p)* What would be your final model for the data? Justify your answer.

```{r, echo=F}
#Mateix error q anterior ...
cat("A model predicting BB_COUNT with all the other variables, to see if they are all significant in predicting the outcome variable ")
```
```{r}
mdl_pois_reg_laboral <- glm(formula = BB_COUNT ~ HIGH_T + LABOR_YESNO + as.double(PRECIP) + as.double(LOW_T) + Date,data = data, family = "poisson"(link="log"))
summary(mdl_pois_reg_laboral)
```
#### *1.q)* Give examples of outcomes that can be modelled using a Poisson regression, such as the number of goals in a handball match.
.

```{r}

```
```{r, echo=F}
cat("Some examples of outcomes would be:
    Number of traffic accidents based on weather conditions.
    Number of visits to E.R. from different causes of injury.
    
    ")
```










